Sterile Neutrinos, a Summary

### 2015-04-15 p3_505/presentation/pres.tex
added MINOS slide
matts changes
small changes
lots of updates to paper

\documentclass[mathserif,18pt,xcolor=table]{beamer}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bbm}
\usepackage{ulem}
\usepackage{slashed}
\usepackage{graphicx}
\usepackage{textpos}
\usepackage{listings}
\usepackage{epsfig}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{fixltx2e}
\definecolor{dukeblue}{RGB}{0,0,156}
\definecolor{dukedarkblue}{RGB}{0,26,87}
\definecolor{dukeblack}{RGB}{79,79,79}
\definecolor{dukegray}{RGB}{79,79,79}
\definecolor{dukesecbrown}{RGB}{217,200,158}
\definecolor{dukesecblue}{RGB}{127,169,174}
\mode<presentation> {
  \usetheme{Boadilla}  
  \setbeamercovered{invisible}
  \setbeamertemplate{navigation symbols}{}  
  \setbeamertemplate{frametitle}[default][center]
  \setbeamerfont{title}{series=\bfseries,parent=structure}
  \setbeamerfont{frametitle}{series=\bfseries,parent=structure}
  \setbeamerfont{subtitle}{size=\scriptsize,series=\bfseries,parent=structure}
  \setbeamerfont{author}{size=\scriptsize,parent=structure}
  \setbeamerfont{institute}{size=\small,series=\bfseries,parent=structure}
  \setbeamerfont{date}{size=\scriptsize,parent=structure}
  \setbeamerfont{footline}{size=\tiny,parent=structure}
  \setbeamercolor{normal text}{bg=white,fg=dukeblack}
  \setbeamercolor{structure}{fg=dukeblue}
  \setbeamercolor{alerted text}{fg=red!85!black}
  \setbeamercolor{item projected}{use=item,fg=black,bg=item.fg!35}
  \setbeamercolor*{palette primary}{use=structure,fg=white, bg=dukeblue}
  \setbeamercolor*{palette secondary}{use=structure,bg=dukedarkblue,fg=white}
  \setbeamercolor*{framesubtitle}{fg=dukegray}
  \setbeamercolor*{block title}{parent=structure,fg=black,bg=dukeblue}
  \setbeamercolor*{block body}{fg=black,bg=dukeblack!10}
  \setbeamercolor*{block title alerted}{parent=alerted text,bg=black!15}
  \setbeamercolor*{block title example}{parent=example text,bg=black!15}
}

@@

\begin{frame}
  \frametitle{Outline}
  \begin{itemize}
  \item neutrinos are kewl
  \item and maybe sterile who knows
  \end{itemize}
\end{frame}

@@

>>>>

\documentclass[mathserif,18pt,xcolor=table]{beamer}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bbm}
\usepackage{ulem}
\usepackage{feynmp-auto}
\usepackage{slashed}
\usepackage{graphicx}
\usepackage{textpos}
\usepackage{listings}
\usepackage{epsfig}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{fixltx2e}
\definecolor{dukeblue}{RGB}{0,0,156}
\definecolor{dukedarkblue}{RGB}{0,26,87}
\definecolor{dukeblack}{RGB}{79,79,79}
\definecolor{dukegray}{RGB}{79,79,79}
\definecolor{dukesecbrown}{RGB}{217,200,158}
\definecolor{dukesecblue}{RGB}{127,169,174}
\mode<presentation> {
  \usetheme{Boadilla}  
  \setbeamercovered{invisible}
  \setbeamertemplate{navigation symbols}{}  
  \setbeamertemplate{frametitle}[default][center]
  \setbeamerfont{title}{series=\bfseries,parent=structure}
  \setbeamerfont{frametitle}{series=\bfseries,parent=structure}
  \setbeamerfont{subtitle}{size=\scriptsize,series=\bfseries,parent=structure}
  \setbeamerfont{author}{size=\scriptsize,parent=structure}
  \setbeamerfont{institute}{size=\small,series=\bfseries,parent=structure}
  \setbeamerfont{date}{size=\scriptsize,parent=structure}
  \setbeamerfont{footline}{size=\tiny,parent=structure}
  \setbeamercolor{normal text}{bg=white,fg=dukeblack}
  \setbeamercolor{structure}{fg=dukeblue}
  \setbeamercolor{alerted text}{fg=red!85!black}
  \setbeamercolor{item projected}{use=item,fg=black,bg=item.fg!35}
  \setbeamercolor*{palette primary}{use=structure,fg=white, bg=dukeblue}
  \setbeamercolor*{palette secondary}{use=structure,bg=dukedarkblue,fg=white}
  \setbeamercolor*{framesubtitle}{fg=dukegray}
  \setbeamercolor*{block title}{parent=structure,fg=black,bg=dukeblue}
  \setbeamercolor*{block body}{fg=black,bg=dukeblack!10}
  \setbeamercolor*{block title alerted}{parent=alerted text,bg=black!15}
  \setbeamercolor*{block title example}{parent=example text,bg=black!15}
}

@@

\begin{frame}
  \frametitle{Outline}
  \begin{itemize}
  \item Introduction to neutrinos
  \item Standard model neutrinos and neutrino oscillations
  \item Sterile neutrinos
  \item Past sterile evidence
  \item Current and future status of sterile analysis.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Little Neutral One}
  \begin{itemize}
  \item No charge, small mass $\sim 0.320 \pm 0.081 \text{ eV}$ (sum of 3 flavors), spin 1/2 fermions.
  \item Interactions mediated only by the weak interaction.
  \item First theorized due to observed $\beta$-decay spectrum. Distribution of $\beta$ particle energy suggests three-body decay.
  \item Existence of antineutrino is indicated from inverse $\beta$-decay.
  \item Helicity: massless particle? \\
    neutrinos: (always) left-handed from experiments.\\
    antineutrinos: (always) right-handed from experiments.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Neutrinos in the Standard Model}
  \begin{itemize}
  \item In the standard model, exactly zero mass;
    There is no mass term in the SM Lagrangian
  \item Neutrinos and antineutrinos are distinct (Dirac particles)
  \item Exactly three neutrinos with lepton number conservation
  \item Couple to the $W$ and the $Z$ only.
  \end{itemize}
  \vspace{1cm}
  \begin{columns}
    \begin{column}{.5\textwidth}
      \begin{center}
        \begin{fmffile}{Znunu}
          \begin{fmfgraph*}(60,50)
            \fmfstraight
            \fmfleftn{i}{1}\fmfrightn{o}{2}
            \fmf{boson}{i1,v1}
            \fmf{fermion}{o1,v1,o2}
            \fmflabel{$Z$}{i1}
            \fmflabel{$\overline{\nu}$}{o1}
            \fmflabel{$\nu$}{o2}
          \end{fmfgraph*}
        \end{fmffile}
      \end{center}
    \end{column}
    \begin{column}{.5\textwidth}
      \begin{center}
        \begin{fmffile}{Wenu}
          \begin{fmfgraph*}(60,50)
            \fmfstraight
            \fmfleftn{i}{1}\fmfrightn{o}{2}
            \fmf{boson}{i1,v1}
            \fmf{fermion}{o1,v1,o2}
            \fmflabel{$W^-$}{i1}
            \fmflabel{$\overline{\nu}$}{o1}
            \fmflabel{$e^-$}{o2}
          \end{fmfgraph*}
        \end{fmffile}
      \end{center}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}
  \frametitle{Neutrinos beyond the Standard Model}
  \begin{itemize}
  \item Neutrino has mass as indicated by oscillations.
  \item Helicity can therefore change, not permanent anymore.
  \item There are many models and extensions added to the Standard Model to resolve questions.
  \item Majorana particle and exact mass? $0\nu2\beta$ experiments are taking this on.
  \item Where does the mass come from? See-Saw model is an idea.
  \item CP violation and lepton number conservation? No if they are Majorana.
  \item Sterile neutrinos, interacting only through gravitation?
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Flavor Oscillations}
  \begin{itemize}
  \item Mass eigenstates and flavor eigenstates are different; a unitary mixing matrix governs the transformation from one basis to the other.
    \begin{align}
      |\nu_j\rangle &= \sum_\alpha U_{\alpha j}|\nu_\alpha\rangle \\
      |\nu_j(t)\rangle\rangle &= e^{-iE_j t}|\nu_j\rangle \\
      |\nu_\alpha\rangle &= \sum_{j=1}U^{*}_{\alpha j}|\nu_j\rangle \\
      \rightarrow |\nu_\alpha(t)\rangle &= \sum_{\beta=e,\mu,\tau}\sum_{j=1,2,3} U_{\beta j}U^*_{\alpha j}e^{-iE_jt}|\nu_\beta\rangle \\
      P(\nu_\alpha\rightarrow \nu_\beta) &= |\langle\nu_\beta|\nu_\alpha\rangle|^2 \\
      &= \sum_{j=1}^3\sum_{k=1}^3 U^*_{\alpha j}U_{\alpha k}U_{\beta j}U_{\beta k}^* e^{-i(E_j-E_k)k}.
    \end{align}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Flavor Oscillations}
  \begin{itemize}
  \item Using $E\approx p$ (ultrarelativistic neutrinos), and $L = ct$:
    \begin{align}
      E_j - E_k &= \left(m_j^2 - m_k^2\right)/2E = \Delta m_{jk}^2/2E \\
      P &= \sum_{j=1}^3\sum_{k=1}^3 U^*_{\alpha j}U_{\alpha k}U_{\beta j}U_{\beta k}^* e^{-i\Delta m_{jk}^2L/2E}
    \end{align}
  \item And therefore a complete expression without hte complex exponential is:
    \begin{align}
        P(\nu_{\alpha}\rightarrow\nu_{\beta}) = \delta_{\alpha\beta} & - 4\sum_{j>k}\Re\left(U_{\alpha j}^*U_{\alpha k}U_{\beta j}U_{\beta k}^*\right)\sin^2\left(\frac{\Delta m_{jk}^2}{4E}L\right) \nonumber \\
  & + 2\sum_{j>k}\Im\left(U_{\alpha j}^*U_{\alpha k}U_{\beta j}U_{\beta k}^*\right)\sin\left(\frac{\Delta m_{jk}^2}{2E}L\right).
    \end{align}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Flavor Oscillations}
  \begin{itemize}
  \item The Pontecorvo Maki, Nakagawa, Sakata Matrix
\begin{align}
  U & =
  \begin{pmatrix}
    U_{e1}    & U_{e2}    & U_{e3}   \\
    U_{\mu1}  & U_{\mu2}  & U_{\mu3} \\
    U_{\tau1} & U_{\tau3} & U_{\tau3}
  \end{pmatrix} \\
  \label{eq:umatrix}
  & = \begin{pmatrix}
    1 & 0 & 0 \\
    0 & c_{23} & s_{23} \\
    0 & -s_{23} & c_{23} \end{pmatrix} \times
  \begin{pmatrix}
    c_{13} & 0 & s_{13}e^{-i\delta} \\
    0      & 1 & 0 \\
    -s_{13}e^{i\delta} & 0 & c_{13} \end{pmatrix} \times
  \begin{pmatrix}
    c_{12} & s_{12} & 0 \\
    -s_{12} & c_{12} & 0 \\
    0 & 0 & 1
  \end{pmatrix} \\
  & = \begin{pmatrix}
    c_{12}c_{13} & s_{12}c_{13} & s_{13}e^{-i\delta} \\
    -s_{12}c_{23} - c_{12}s_{23}s_{13}e^{i\delta} & c_{12}c_{23} - s_{12}s_{23}s_{13}e^{i\delta} & s_{23}c_{13} \\
    s_{12}s_{23} - c_{12}c_{23}s_{13}e^{i\delta} & -c_{12}s_{23} - s_{12}c_{23}s_{13}e^{i\delta} & c_{23}c_{13}
  \end{pmatrix}.
\end{align}
\end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Flavor Oscillations}
  \begin{itemize}
    \item Simple two neutrino oscillation model.
    \begin{align}
      \begin{split}
        \begin{pmatrix}
          |\nu_\alpha\rangle \\
          |\nu_\beta\rangle
        \end{pmatrix} = 
        \begin{pmatrix}
          U_{\alpha 1} & U_{\alpha 2} \\
          U_{\beta  1} & U_{\beta  2}
        \end{pmatrix}
        \begin{pmatrix}
          |\nu_1\rangle \\
          |\nu_2\rangle
        \end{pmatrix}
      \end{split}
    \end{align}
  \item Probability of oscillation from $\nu_\alpha$ to $\nu_\beta$:
    \begin{align}
      P(\nu_\alpha \rightarrow \nu_\beta) = \left|\langle \nu_\beta(x,t)|\nu_\alpha(0,0\rangle\right|^2
    \end{align}
  \item In this simple two neutrino model this yields probability expressions for the form:
    \begin{align}
      P(\nu_\mu\rightarrow\nu_e)   &= \sin^2(2\theta) \sin^2\left(\frac{1.27\Delta m^2[\text{eV}^2] L[\text{km}]}{E[\text{GeV}]}\right) \\
      P(\nu_\mu\rightarrow\nu_\mu) &= 1 - \sin^2(2\theta) \sin^2\left(\frac{1.27\Delta m^2[\text{eV}^2] L[\text{km}]}{E[\text{GeV}]}\right)
    \end{align}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Flavor Oscillations}
  \begin{center}
    \includegraphics[width=.6\linewidth]{../figures/oscillation_prob.pdf}    
  \end{center}
  \begin{itemize}
\item Oscillation and survival probabilities for the two neutrino state with an initial $\nu_{\mu}$. The values for $\Delta m^2$ and $\sin^22\theta$ ($2.41\times 10^{-3}$ eV$^2$ and 0.97, respectively) are taken from 2013 MINOS result (PRL, 110, 251801)    
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{See-Saw Model}
  \begin{itemize}
  \item Popular way to explain why there are light left-handed neutrinos and no observed right-handed neutrinos.
  \item Right-handed states would be sterile -- don't interact (or interact only very weakly) via weak weak interactions.
  \end{itemize}
  \begin{center}
    \includegraphics[width=.75\linewidth]{../figures/seesaw.pdf}
  \end{center}
\end{frame}

\begin{frame}
  \frametitle{See-Saw continued}
  \begin{align}
    \begin{split}
    \begin{pmatrix}
      \overline{\nu} & \overline{N}
    \end{pmatrix}
    \begin{pmatrix}
      0 & m_D \\
      m_D & M
    \end{pmatrix}
    \begin{pmatrix}
      \nu \\ N
    \end{pmatrix}
    \end{split}
  \end{align}
  \begin{itemize}
    \item For the Dirac mass much ness than $M$, eigenvalues are $-m_D^2/M$ and $M$, with eigenstates made from mostly $\nu$ and $N$, respectively.
    \item This means that as $M$ goes up or down $m_D^2/M$ does the opposite -- hence the See-Saw.
    \item We are left with light-left handed neutrinos and heavy right-handed.
    \item There are 3 types of See-Saw model (I, II, III) that involve the introduction of isospin singlets, scalar triplets, and fermion triplets.
    \item Drawing distinctions is beyond the scope of this project.
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{The MINOS/MINOS+ Experiment}
  \begin{itemize}
  \item Long baseline, disappearance experiment
  \item Latest results now exclude much of the LSND allowed phase space
    \begin{center}
      \includegraphics[width=.9\linewidth]{../figures/minos2.pdf}
    \end{center}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{The Super-Kamiokande Experiment}
  \begin{itemize}
  \item Atmospheric, TODO experiment
  \item Latest results TODO
    \begin{center}
      \includegraphics[width=.5\linewidth]{../figures/sk1.pdf}
    \end{center}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{The Daya Bay Experiment}
  \begin{itemize}
  \item Reactor, TODO experiment
  \item Latest results TODO
    \begin{center}
      \includegraphics[width=.5\linewidth]{../figures/daya1.pdf}
    \end{center}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Conclusions}
  \begin{itemize}
  \item Sterile neutrinos\ldots
  \end{itemize}
\end{frame}
